The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-19 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions as discussed (small changes in figures and captions, tables 3 and 4, formula for beta, consistent use of unifying terms for scenarios, structure of section 5, and some text revisions in sections 4 and 5), still to do: further text revision in section 5
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2575 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: cc4b21dc7789b8a0c13678111c8619bed8200665

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref}
\usepackage[all]{hypcap}
\usepackage{enumitem}
\usepackage{color,soul}

@@

Each generated data set consists of the response and regressor variables, one true split 
variable and nine noise split variables as listed in Table~\ref{tab:variables} 
together with the corresponding distributions.

\begin{table}
\centering
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\caption{Variables included in the data generating process of 
the first scenario (stump).
\label{tab:variables}}
\end{table}

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$
can depend on the true split variable $Z_1$. More specifically, three different scenarios
are considered for the coefficients $\beta_0$ and $\beta_1$:
%
\begin{enumerate}
  \item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
  \item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
        (at $0$).
  \item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}
%
See Figure~\ref{fig:dgp_stump} for an illustration. More precisely, the coefficient that
changes ($\beta_j$ with $j\in\{0,1\}$) switches between two values at the split point $\xi$:
%
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{j}  \quad \text{if } Z_1 < \xi \\
+\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \geq \xi
\end{cases}
\end{align*}
%
%while in the continuous case a simple linear function is considered in the form of
%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.

@@

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$
and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three cases: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

For the second simulation scenario the same basic setting with the same variables is used
with an additional split in $Z_2$ resulting in a tree with three terminal nodes. The first
split (in $Z_2$, at~$\xi$) induces a change in the slope parameter $\beta_1$ while the second
split (in $Z_1$, also at~$\xi$) corresponds to a change in the intercept $\beta_0$.
The underlying tree structure for the data generating process is illustrated in 
Figure~\ref{fig:dgp_tree}.

@@

\begin{figure}%[t]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
plot(p, edge_panel = ep, terminal_panel = tp,
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(-1,-1), col = pal_dgp[2], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[3], lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha(pal_dgp[2], 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha(pal_dgp[3], 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c(pal_dgp[1], pal_dgp[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
@
\endminipage
\caption{\label{fig:dgp_tree}Left panel: True tree structure for the second simulation scenario 
with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$.}
\end{center}
\end{figure}

\subsection{Evaluation}
\label{sec:evaluation}
The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
on 100 replications per step each consisting of 250 observations. 
To compare the performance of the evaluated testing strategies for each step 
in the first scenario (stump with one split) the following criteria are considered:
the $p$-values pertaining only to the true split variable $Z_1$;
and the proportion of replications for which the $p$-value of $Z_1$
is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
selected for splitting in a pre-pruning approach).
For the second scenario (tree with two splits) the adjusted Rand index~(ARI) is calculated 
as a measure of similarity between the true tree structure and the fitted model tree.

@@

@@

\subsection[Residuals vs. full model scores]{Residuals vs.\ full model scores}
\label{sec:scores}

To assess the impact of using only residuals or the full score matrix four testing
strategies are compared: CTree, GUIDE, MOB -- all three using the default specification
as shown in Table~\ref{tab:buildingblocks} -- and additionally the GUIDE strategy using
the full scores instead (denoted by GUIDE+scores).

@@

               strip =strip.custom(factor.levels = c(expression("varying"~beta[0]~"and"~beta[1]), 
                                                     expression("varying"~beta[0]), 
                                                     expression("varying"~beta[1])),
                                   bg = "gray90"),
               strip.left=strip.custom(factor.levels = c(expression(xi == 0),                                                              expression(xi == 0.8)),
                                   bg = "gray90"))

@
\caption{\label{fig:scores_stump}
CTree, MOB, GUIDE, and GUIDE+scores evaluated on the first simulation scenario (stump with one split). 
Their performances are compared based on the proportion of replications where the true split 
variable $Z_1$ is selected over increasing effect size $\delta$ (100 replications of 
250 observations per step), once for a true split point $\xi = 0$ (top row) and once 
for $\xi = 0.8$ (bottom row). 
Left: varying intercept $\beta_0$. Middle: varying slope parameter $\beta_1$.
Right: both coefficients varying.}
\end{figure}

Figure~\ref{fig:scores_stump} shows that overall CTree and MOB (both using full 
model scores) perform clearly better than the original GUIDE version (using residuals 
only) as they accomplish to select the true split variable $Z_1$ in a higher proportion
of replications. Even though the advantage of CTree and MOB evens out in certain situations
such as for a high effect size (see, e.g., for $\xi=0$ and varying $\beta_0$), GUIDE is 
never ahead of them in any of the considered settings. Moreover, the positive effect of 
including all available scores can be observed particularly well in the top middle panel 
where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to 
GUIDE. The fact that residual-based methods fail to detect changes that do not effect the
expected value has already been elaborated in the literature of structural change
analysis \citep[see, e.g.,][]{Ploberger+Kraemer:1992} and is also discussed in 
Section~\ref{sec:discussion}. Moreover, the advantage of including full models scores 
clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
to GUIDE+scores in case of a fixed slope parameter (left panels).

For a central split point ($\xi=0$, upper panels) all strategies perform similarly
well as long as the intercept varies (left and right panel) with CTree and MOB 
being only slightly ahead. However, for the split point being close to the upper 
edge of the considered interval ($\xi=0.8$, lower panels) the performance of all of 
the applied strategies decreases. 
While both GUIDE versions struggle to detect the correct split variable even for a 
high effect size, MOB is clearly ahead with CTree leading to the second best results.
(In order to show that the here observed advantage of MOB over CTree is mainly due to the 
abrupt shift in the model parameters $\beta_0$ and $\beta_1$ an additional evaluation is 
presented in Appendix~\ref{app:contbeta} where these parameters change continuously.)
From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
can be deduced that categorizing split variables and binarizing scores has a negative effect 
on the ability to detect the right split variable. To elaborate these effects more in detail the results for all possible combinations of the building blocks are discussed in a full factorial 3-way analysis in the following section.

@@

\subsection{Full factorial analysis}\label{sec:3way}
To investigate the effects of each of the building blocks separately the most general 
case of the first simulation scenario (stump with one split) where the intercept and 
slope parameter are both varying is considered. In the underlying evaluation all 
possible combinations of the building blocks have been included. The different levels of 
each of the three building blocks are listed in Table~\ref{tab:levels}. A more detailed
overview of all 12 combinations can be found in Appendix~\ref{app:combinations}, 
Table~\ref{tab:combinations_appendix}.

\begin{table}[h!]
\centering
\begin{tabular}{ l l l }
\hline
Building block         & Levels              & Effect \\
\hline
Residuals vs.\ scores  & \textit{residuals}  & Use only residuals\\
                       & \textit{scores}     & Use full score matrix\\
Categorization         & \textit{cat}        & Categorize split variables\\
                       & \textit{max}        & Selected split variables maximally w.o. categorization\\
                       & \textit{lin}        & Selected split variables linearly w.o. categorization\\
Binarization           & \textit{bin}        & Binarize scores\\
                       & \textit{lin}        & Use residuals/score values without binarization\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing residuals/scores decreases the performance 
as it leads to higher $p$-values corresponding to the true split variable $Z_1$ 
and thus to lower power in the presented settings. 
For a central true split point (left panel) this effect remains at an almost constant level 
regardless of whether possible split variables are categorized or not. However, in case 
of the true split point being closer to the upper margin ($\xi=0.8$, right panel), categorizing
split variables increases the $p$-value of $Z_1$ even more. 
Here, the maximal selection performs best while the linear selection shows its disadvantage in 
detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximal
selection of a split variable is advantageous. But unless one of the breakpoints used for binning possible split variables is close to the true split point (as for $\xi=0$, left panel),
this categorization weakens the performance of a testing strategy.

>>>>

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref,dsfont}
\usepackage[all]{hypcap}
\usepackage{enumitem}
\usepackage{color,soul}

@@

\subsubsection{``Stump'' scenario}
Each generated data set consists of the response and regressor variables, one true split 
variable and nine noise split variables as listed in Table~\ref{tab:variables} 
together with the corresponding distributions.

\begin{table}
\centering
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$\\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ (alternating) \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\caption{\label{tab:variables}Variables included in the data generating process 
as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a split variable, 
not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_i(Z_1,Z_2)$, $i=1,2$.}
\end{table}

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$
can depend on the true split variable $Z_1$. More specifically, three different variations
are considered for the coefficients $\beta_0$ and $\beta_1$:
%
\begin{enumerate}
  \item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
  \item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
        (at $0$).
  \item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}
%
See Figure~\ref{fig:dgp_stump} for an illustration. More precisely, the coefficient that
changes ($\beta_j$ with $j\in\{0,1\}$) switches between two values at the split point $\xi$:
%
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \leq \xi \\
+\delta \cdot (-1)^{j}  \quad \text{if } Z_1 > \xi
\end{cases}
\end{align*}
%
%while in the continuous case a simple linear function is considered in the form of
%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.

@@

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure applied in the data generating process for the ``stump'' scenario with the true model parameters $\beta_0$
and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three cases: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

\subsubsection{``Tree'' scenario}
For the second simulation scenario (``tree'' scenario) the same basic setting with the same 
variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
However, not only $Z_1$ but also $Z_2$ is used as a split variable. Therefore, an additional split
is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
change in the slope parameter $\beta_1$ while the second split (in $Z_1$, also at~$\xi$) corresponds 
to a change in the intercept $\beta_0$.
Hence, contrary to the ``stump'' scenario where three variations are considered for the parameters 
$\beta_0$ and $\beta_1$, only this one variation with both parameters varying is investigated for 
the ``tree'' scenario. In particular, the parameters depend on the split variables in the following way:
%
\begin{align*}
\beta_{0}(Z_1, Z_2) = \begin{cases}
0       \qquad \text{if } Z_2 \leq \xi \\
-\delta  \quad \text{ if } Z_1 \leq \xi \land Z_2 >\xi \\
+\delta  \quad \text{ if } Z_1 > \xi \land Z_2 >\xi
\end{cases}
\end{align*}
\begin{align*}
\beta_{1}(Z_1, Z_2) = \beta_{1}(Z_2)= \begin{cases}
+\delta \quad \text{if } Z_2 \leq \xi \\
-\delta \quad \text{if } Z_2 > \xi
\end{cases}
\end{align*}
%
The underlying tree structure for the data generating process is illustrated in 
Figure~\ref{fig:dgp_tree}.

@@

\begin{figure}%[t]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
plot(p, edge_panel = ep, terminal_panel = tp,
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(-1,-1), col = pal_dgp[2], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[3], lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha(pal_dgp[2], 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha(pal_dgp[3], 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c(pal_dgp[1], pal_dgp[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
@
\endminipage
\caption{\label{fig:dgp_tree}Left panel: True tree structure for the ``tree'' scenario 
with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$.}
\end{center}
\end{figure}

\subsection{Evaluation}
\label{sec:evaluation}
The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
on 100 replications per step each consisting of 250 observations. 
To compare the performance of the evaluated testing strategies for each step 
in the ``stump'' scenario the following criteria are considered:
the $p$-values pertaining only to the true split variable $Z_1$;
and the proportion of replications for which the $p$-value of $Z_1$
is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
selected for splitting in a pre-pruning approach), denoted by the 
``selection probability''.
For the ``tree'' scenario the adjusted Rand index~(ARI) is calculated 
as a measure of similarity between the true tree structure and the fitted model tree.

@@

In the following we first investigate the properties of the testing
strategies in the ``stump'' scenario, focusing on the tests' $p$-values
for the true split variable $Z_1$ and its corresponding selection
probability (i.e., $Z_1$ being significant at 5\% level and having the
lowest $p$-value among all split variables). Section~\ref{sec:scores} begins by
highlighting the importance of using full model scores vs.\ residuals only
before Section~\ref{sec:3way} considers all building blocks (scores vs.\ residuals,
binarization of these, and categorization of the split variable).
Subsequently, the ``tree'' scenario is employed to investigate how the
performance of the tests affects growing the trees overall. This is
evaluated using the adjusted Rand index for trees grown by pre-pruning and 
cost-complexity post-pruning (Section~\ref{sec:tree}).

@@

\subsection[``Stump'' scenario: Residuals vs. full model scores]{``Stump'' scenario: Residuals vs.\ full model scores}
\label{sec:scores}

A crucial difference between the testing strategies of CTree/MOB and GUIDE
is the difference between using only the residuals vs. the full model
scores. In the literature on structural change tests there is evidence
that residual-based tests can only capture parameter differences that
affect the conditional mean \citep[see e.g.,][]{Ploberger+Kraemer:1992}.
Hence we compare the CTree, MOB, and GUIDE algorithms 
-- all three using the default specification as shown in Table~\ref{tab:buildingblocks} -- 
and additionally consider a new GUIDE flavor, denoted GUIDE+scores, that uses
binarized scores rather than residuals.

%To assess the impact of using only residuals or the full score matrix four testing
%strategies are compared: CTree, GUIDE, MOB -- all three using the default specification
%as shown in Table~\ref{tab:buildingblocks} -- and additionally the GUIDE strategy using
%the full scores instead (denoted by GUIDE+scores).

@@

               strip =strip.custom(factor.levels = c(expression("varying"~beta[0]~"and"~beta[1]), 
                                                     expression("varying"~beta[0]), 
                                                     expression("varying"~beta[1])),
                                   bg = "gray90"),
               strip.left=strip.custom(factor.levels = c(expression(xi == 0~" (50%)"),                                                              expression(xi == 0.8~" (90%)" )),
                                   bg = "gray90"))

@
\caption{\label{fig:scores_stump}
Selection probability of the true split variable $Z_1$ for
testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the ``stump''
scenario. Probabilities are estimated over increasing effect size $\delta$
with 100 replications of 250 observations per step. The true split point
in $Z_1$ is either the median 0 (top) or the 90\% quantile 0.8 (bottom)
for either varying intercept $\beta_0$ (left), varying slope $\beta_1$
(middle), or both coefficients varying (right).}
\end{figure}

Figure~\ref{fig:scores_stump} shows that overall CTree and MOB (both using full 
model scores) perform clearly better than the original GUIDE version (using residuals 
only) as they accomplish to select the true split variable $Z_1$ in a higher proportion
of replications. Even though the advantage of CTree and MOB evens out in certain situations
such as for a high effect size (see, e.g., for the median $\xi=0$ and varying $\beta_0$), GUIDE is 
never ahead of them in any of the considered settings. Moreover, the positive effect of 
including all available scores can be observed particularly well in the top middle panel 
where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to 
GUIDE. 
%The fact that residual-based methods fail to detect changes that do not effect the
%expected value has already been elaborated in the literature of structural change
%analysis \citep[see, e.g.,][]{Ploberger+Kraemer:1992} and is also discussed in 
%Section~\ref{sec:discussion}. 
Furthermore, the advantage of including full models scores 
clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
to GUIDE+scores in case of a fixed slope parameter (left panels).

For a central split point ($\xi=0$ corresponding to the median, upper panels) all strategies 
perform similarly well as long as the intercept varies (left and right panel) with CTree and MOB 
being only slightly ahead. However, for the split point being close to the upper 
edge of the considered interval ($\xi=0.8$ corresponding to the $90$\%-quantile, lower panels) 
the performance of all of the applied strategies decreases. 
While both GUIDE versions struggle to detect the correct split variable even for a 
high effect size, MOB is clearly ahead with CTree leading to the second best results.
(In order to show that the here observed advantage of MOB over CTree is mainly due to the 
abrupt shift in the model parameters $\beta_0$ and $\beta_1$ an additional evaluation is 
presented in Appendix~\ref{app:contbeta} where these parameters change continuously.)
From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
can be deduced that categorizing split variables and binarizing scores has a negative effect 
on the ability to detect the right split variable. To elaborate these effects more in detail the results for all possible combinations of the building blocks are discussed in a full factorial 3-way analysis in the following section.

@@

\subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
To investigate the effects of each of the building blocks separately the most general 
case of the ``stump'' scenario where the intercept and 
slope parameter are both varying is considered. In the underlying evaluation all 
possible combinations of the building blocks have been included. The different levels of 
each of the three building blocks are listed in Table~\ref{tab:levels}. A more detailed
overview of all 12 combinations can be found in Appendix~\ref{app:combinations}, 
Table~\ref{tab:combinations_appendix}.

\begin{table}[h!]
\centering
\begin{tabular}{ l l l }
\hline
Building block               & Levels              & Transformation \\
\hline
Residuals vs.\ scores        & \textit{residuals}  & $r(Y,X,\hat{\beta})$\\
\quad ($h$ transformation)   & \textit{scores}     & $s(Y,X,\hat{\beta})$\\
\hline
Binarization                 & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
\quad (of $h$)               & \textit{no}         & $h$ without binarization\\
\hline
Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)$\\
\quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
                             & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)$\\
                             &                     & \quad indicating the crossed quantile\\
                             & \textit{lin}        & $g(Z)=Z$\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing residuals/scores decreases the performance 
as it leads to higher $p$-values corresponding to the true split variable $Z_1$ 
and thus to lower power in the presented settings. 
For a central true split point (left panel) this effect remains at an almost constant level 
regardless of whether possible split variables are categorized or not. However, in case 
of the true split point being closer to the upper margin ($90$\%-quantile $\xi=0.8$, right panel),
categorizing split variables increases the $p$-value of $Z_1$ even more. 
Here, the maximal selection performs best while the linear selection shows its disadvantage in 
detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximal
selection of a split variable is advantageous. But unless one of the breakpoints used for binning possible split variables is close to the true split point (as for the median $\xi=0$, left panel),
this categorization weakens the performance of a testing strategy.
